---
title: "Supplement. Quantifying natural disturbances using a large-scale dendrochronological reconstruction to guide forest management"
author: "Cada V., Trotsiuk V., ...."
date: "2/8/2019"
output: 
  word_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Libraries and settings

```{r libraries, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(dplyr) 
library(tibble)
library(tidyr)
library(zoo)
library(pracma)
library(readr)
```

# Functions

```{r functions, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

priorGrowth <- function(x, windowLength = 10){
  rollapply( x, 
             width = windowLength,
             FUN = mean,
             fill = NA,
             align = "right",
             na.rm = T,
             partial = TRUE)
}

followGrowth <- function(x, windowLength = 10){
  rollapply( lead(x, 1), 
             width = windowLength,
             FUN = mean,
             fill = NA,
             align = "left",
             na.rm = T,
             partial = TRUE)
}

peakDetection <- function(x, threshold, mindist = 20, nups = 2){
  #' @description identify the index of year when release event occur
  #' @param x a vector of absolute increase change
  #' @param threshold a minimum ai value in mm
  #' @param mindist  minimum distance between two consecutive peaks in years
  #' @param nups number of increasing steps before the peak
  
  x <- ifelse(is.na(x), -0.2, x)
  
  x <- findpeaks(x, 
                 minpeakheight = threshold,
                 minpeakdistance = mindist,
                 nups = nups) 
  
  if(is.null(x)){
    NA
  }else{
    matrix(x, ncol = 4)[,2]
  }
}

keepRelease <- function(year, type, n = 20){
  #' @description calculate the distance between gap origin and releases
  #' @param year the vector of years for event
  #' @param type type of the event (release or gap)
  #' @param n number of years to be checked
  
  keep <- rep('yes', length(year))
  
  if(any(type %in% 'gap')){
    diffyear <- year - year[type %in% 'gap']
    keep[diffyear < n & type %in% 'release'] <- 'no'
  }
  keep
}

growthCalculate <- function(data = data, windowLength = 10){
  #' @description take the list of data prepared by 'dist_get_data' function and calculate the growth change, plus age and dbh of the trees
  #' @param data a list of tree tables
  #' @param windowLength the length of the window for ai calculation
  
  # data quality check
  options(error = NULL) # not to enter debug mode
  
  # perform the checks
  if(!is.list(data)) stop('The input data is not a list of three tables')
  if(!identical(c('core',"dist_param","ring"), ls(data))) stop('The input data tables dont match with required')
  
  # calculate the age, dbh, and the growth change
  inner_join(
    data$ring,
    data$core,
    by = 'core_id'
  ) %>%
    arrange(core_id, year) %>%
    group_by(core_id) %>%
    mutate(dbh_growth = ifelse(row_number() == 1, incr_mm + missing_mm, incr_mm),
           dbh_growth = cumsum(dbh_growth) * 2,
           dbh_mm = ifelse(is.na(dbh_mm), max(dbh_growth), dbh_mm),
           dbh_coef = max(dbh_mm) / max(dbh_growth),
           dbh_growth = dbh_growth * dbh_coef,
           age = year - min(year) + missing_years + 1,
           pg = priorGrowth(incr_mm, windowLength = windowLength),
           fg = followGrowth(incr_mm, windowLength = windowLength),
           ai = fg - pg) %>%
    select(dist_param, tree_id, core_id, ring_id, year, incr_mm, age, dbh_mm = dbh_growth, ai, fg, pg) ->
    data$ring
  
  return(data)
}

releaseCalculate <- function(data = NULL,  gapAge = c(5:15), nprol = 7){
  #' @description function calculate the releases for individual trees
  #' @param data a list of three dataframes, output of growthCalculate function
  #' @param nprol number of years to consider that release is sustaind
  #' @param gapAge age of the tree when it shall be tested for gap origin
  
  # data quality check
  options(error = NULL) # not to enter debug mode
  
  # perform the checks
  if(!is.list(data)) stop('The input data is not a list of three tables')
  if(!identical(c('core',"dist_param","ring"), ls(data))) stop('The input data tables dont match with required')
  
  aith <- data$dist_param  %>% select(dist_param, ai_mm) %>% deframe()
  gapth <- data$dist_param  %>% select(dist_param, gap_mm) %>% deframe()
  
  # calculate releases
  data$ring %>%
    arrange(year) %>%
    group_by(core_id) %>%
    mutate(event = ifelse(row_number() %in%  peakDetection(x = ai, threshold = aith[first(as.character(dist_param))], nups = 1,  mindist = 30), 'release', NA),
           event = ifelse(lead(fg, nprol) <= pg, NA, event),
           event = ifelse(lag(pg, nprol) >= fg, NA, event)) %>%
    filter(!is.na(event)) %>%
    select(core_id, year, event) ->
    release.event
  
  # calculate the gap origin 
  data$ring %>% 
    filter(age %in% gapAge) %>%
    arrange(year) %>%
    group_by(core_id) %>%
    summarise(dist_param = first(dist_param),
              gapGrowth = mean(incr_mm, na.rm = T),
              N = n(),
              year = min(year)) %>%
    filter(N >= 5,
           gapGrowth >= gapth[as.character(dist_param)]) %>%
    mutate(event = 'gap') %>%
    select(core_id, year, event) ->
    gap.event
  
  # add those that don't have any event
  data$ring %>%
    filter(!core_id %in% c(unique(gap.event$core_id), unique(release.event$core_id))) %>%
    group_by(core_id) %>%
    summarise(year = min(year)) %>%
    mutate(event = 'no event') ->
    no.event
  
  # add together the events
  bind_rows(release.event, gap.event, no.event) %>%
    arrange(year) %>%
    group_by(core_id) %>%
    mutate(keeprel = keepRelease(year, event, n = 30)) %>%
    ungroup() %>%
    filter(keeprel %in% 'yes') %>%
    inner_join(., data$ring, by = c('core_id', 'year')) %>%
    select(ring_id, dist_param, year, age, dbh_mm, ai, event) ->
    data$event
  
  return(data)
  
}

mdsFun <- function(ca, k = 30, bw = 5, st = 7){
  #' @description return a vector of the fited KDE function
  #' @param ca arranged vector of the canopy area values
  #' @param k a windows length, default 30
  #' @param bw a smoothing bandwidth to be used, default = 5
  #' @param st a standartization value, to scale back to canopy area
  
  rollapply( ca, 
             width = k,
             FUN = function(x){n <- length(x); density(1:n, weights = x, bw = bw, n = n)$y[round((n+1)/2)]* 100/st},
             fill = 0,
             align = "center",
             partial = TRUE)
}
```

# Data

```{r data, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
plots_clean <- read_csv("plots_clean_app.csv")

dist_patches <- read_csv("dist_patches_app.csv") %>% filter(stand_size > 20)

data_list <- list(
  dist_param = read_csv("dist_param_app.csv"),
  core = read_csv("core_app.csv"),
  ring = read_csv("ring_app.csv")
)

data_all <- read_csv("data_all_app.csv")

```

# Calculate the growth change and releases at the tree level

```{r growth change, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
data.growth <- growthCalculate(data = data_list, windowLength = 10)

data.release <- releaseCalculate(data = data.growth, gapAge = c(5:15), nprol = 7)
```

# Calculate the Kernel density and find peaks at the plot level

## Disturbance history data

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

data_all %>% 
  rowwise() %>%
  mutate(ca = eval(parse(text = dbh_ca_f))) %>% 
  ungroup() %>%
  do({
    x <- .
    inner_join(
      x %>% group_by(plotid, species, event, year) %>% summarise(ca = sum(ca)),
      x %>% distinct(tree_id, .keep_all = T) %>% group_by(plotid) %>% summarise(ca_f = sum(ca), n = n()) %>% filter(n >= 5),
      by = 'plotid'
    ) 
  }) %>%
  ungroup() %>%
  mutate(ca = ca * 100 / ca_f) %>%
  arrange(plotid, year) %>%
  filter(year %in% c(1600:2010)) %>%
  select(plotid, species, event, year, ca) %>%
  gather(plot_type, value, -year, -ca, -plotid) %>%
  mutate(plot_type = factor(plot_type, levels = c('species', 'event')),
         Species = factor(value, levels = c('Picea', 'Fagus', 'Abies', 'Acer', 'Pinus', 'Others', 'gap', 'release', 'no event')))->
  data.dist

```

## Calculate the Kernel density (MDS) and the moving sums

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

data.dist %>%
  filter(plot_type == 'species') %>%
  group_by(plotid, year) %>%
  summarise(ca = sum(ca)) %>%
  group_by(plotid) %>%
  complete(year = 1600:2030, fill = list(ca = 0)) %>%
  mutate(value = mdsFun(ca, k = 30, bw = 5, st = 7),
         ca = round(ca, digits = 2),
         severity = rollapply(ca, width = 11, FUN = sum, fill = 0)) %>%
  ungroup() %>%
  filter(year %in% c(1600:2010)) ->
  data.mds

```

## Detect the peaks in Kernel density

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

data.mds %>%
  group_by(plotid) %>%
  filter(row_number() %in% peakDetection(x = value, threshold = 10, nups = 5,  mindist = 10)) %>% 
  mutate(method = '10_10_5') %>%
  ungroup() ->
  data.peaks

```

# Calculate the Kernel density and find peaks at the stand level

## Data

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

data.peaks %>% 
  filter(severity > 10) %>%
  mutate(value = round(value, digits = 5)) %>%
  inner_join(., plots_clean, by = "plotid") %>%
  select(plotid, Xjtsk, Yjtsk, country, newstand, year, ca_per = ca, kde = value, severity) ->
  dist_eventsb

```

## Calculate and bootstrap density function for chronologies of plot level events

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

dist_eventsb %>% 
  select(country, newstand, plotid, year) %>%
  filter(!is.na(newstand)) %>%
  group_by(country, newstand) %>% 
  mutate(nplots = length(unique(plotid))) %>%
  filter(year %in% c(1811:1989)) %>% 
  slice(rep(1:n(), each = 1000)) %>%
  ungroup() %>% 
  mutate(rep = rep(1:1000, times = nrow(.) / 1000)) ->
  standlevel_densityf_boot

standlevel_densityf_boot %>%
  distinct(., country, newstand, plotid, rep) %>% 
  group_by(country, newstand, rep) %>%
  sample_n(., size = 10, replace = TRUE) %>%
  left_join(., standlevel_densityf_boot, by = c("country", "newstand", "plotid", "rep")) %>%
  group_by(country, newstand, year, rep) %>%
  summarise(nevents = n(),
            nplots = mean(nplots),
            freq = nevents / nplots) %>%
  group_by(country, newstand, rep) %>%
  complete(year = c(1780:2020), fill = list(freq = 0)) %>%
  mutate(density_pre = mdsFun(freq, k = 30, bw = 5, st = 7),
         density = rollapply(density_pre, width = 5, FUN = mean, fill = 0)) %>% 
  ungroup() ->
  standlevel_densityf_boot

```

## Find peaks in all (bootstrapped) density functions and select most frequent peaks

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
standlevel_densityf_boot %>%
  filter(dplyr::row_number() %in% peakDetection(x = density, threshold = 0.00001, nups = 5,  mindist = 10),
         year %in% c(1812:1988)) %>%
  group_by(country, newstand, year) %>%
  summarise(freq = n()/100) %>%
  group_by(country, newstand) %>%
  complete(year = c(1780:2020), fill = list(freq = 0)) %>%
  mutate(freqsmooth = mdsFun(freq, k = 11, bw = 1, st = 7)/10) %>%
  filter(row_number() %in% peakDetection(x = freqsmooth, threshold = 0.00001, nups = 0,  mindist = 10)) %>%
  unite(peakid, c("country", "newstand", "year"), sep = "-", remove = FALSE) %>%
  select(peakid, country, newstand, year, freqsmooth) %>%
  ungroup() -> 
  standlevel_peaks
```

## Group plot level events (join plot level events with closest stand level peaks)

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
dist_events_dt <- data.table::data.table(subset(subset(dist_eventsb, !is.na(newstand)), year %in% c(1811:1989)),
  key = c("country", "newstand", "year"))

standlevel_peaks_dt <- data.table::data.table(standlevel_peaks, key = c("country", "newstand", "year"))

dist_eventsb %>%
  left_join(.,
    data.frame(standlevel_peaks_dt[dist_events_dt,
      list(country, newstand, year, peakid, plotid),
      roll = "nearest"]) %>%
      mutate(peakyear = as.numeric(substr(peakid, nchar(peakid) - 3, nchar(peakid)))) %>% 
      select(plotid, year, peakid, peakyear), 
    by = c("plotid", "year")) %>%
  select(plotid, X = Xjtsk, Y = Yjtsk, country, newstand, year, ca_per, kde, peakid, peakyear, severity) ->
  standlevel_dist_events_joined
```

# Calculate the disturbance characteristics and their rotation periods

## Stand proportion disturbed

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
standlevel_dist_events_joined %>% 
  filter(!is.na(peakyear)) %>%
  group_by(newstand, peakyear) %>%
  summarize(nplots_disturb = n()) %>%
  left_join(.,
            plots_clean %>% 
              filter(!is.na(newstand)) %>%
              group_by(locality, country, landscape, newstand) %>%
              summarize(nplots = n()),
            by = "newstand") %>%
  mutate(plotsprop_disturb = nplots_disturb/nplots) ->
  plotsprop_disturb
```

## Rotation period of disturbance severity

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
standlevel_dist_events_joined %>% 
  group_by(country, newstand, plotid) %>% 
  summarize(nevents = n()) %>%
  slice(rep(1:n(), each = 1000)) %>%
  ungroup() %>% 
  mutate(rep = rep(1:1000, times = nrow(.) / 1000)) %>%
  group_by(rep) %>%
  sample_n(., size = 50, replace = TRUE) %>%
  left_join(., standlevel_dist_events_joined, by = c("country", "newstand", "plotid")) %>%
  mutate(severityclass = floor(severity/5) * 5) %>%
  group_by(plotid, rep) %>% 
  mutate(ch_length = 1990 - min(year)) ->
  rotation_severity_boot

rotation_severity_boot %>% 
  group_by(rep, plotid) %>% 
  summarize(ch_length = mean(ch_length)) %>%
  group_by(rep) %>% 
  summarize(rep_length = sum(ch_length)) ->
  rep_length

rotation_severity_boot %>% 
  group_by(rep, severityclass) %>%
  summarize(distevents_n = n()) %>%
  group_by(rep) %>%
  complete(severityclass = seq(10, 150, 5), fill = list(distevents_n = 0)) %>%
  left_join(., rep_length, by = "rep") %>%
  arrange(rep, desc(severityclass)) %>%
  group_by(rep) %>%
  mutate(rotation_s = rep_length/cumsum(distevents_n)) %>%
  group_by(severityclass) %>% 
  summarize(conf95 = quantile(rotation_s, probs = 0.975), 
            conf05 = quantile(rotation_s, probs = 0.025)) %>%
  filter(severityclass < 100) ->
  rotation_severity_confidence

standlevel_dist_events_joined %>%
  left_join(., plots_clean, by = "plotid") %>%
  group_by(landscape, plotid) %>% 
  summarize(ch_length = 1990 - min(year)) %>%
  group_by(landscape) %>% 
  summarize(length = sum(ch_length)) ->
  length_landscapes

standlevel_dist_events_joined %>% 
  select(plotid, year, severity) %>% 
  mutate(severityclass = floor(severity/5) * 5) %>%
  left_join(., plots_clean, by = "plotid") %>%
  group_by(landscape, severityclass) %>%
  summarize(distevents_n = n()) %>%
  arrange(landscape, desc(severityclass)) %>%
  left_join(., length_landscapes, by = "landscape") %>%
  mutate(rotation_s = length/cumsum(distevents_n)) %>%
  select(-distevents_n) %>% 
  filter(severityclass < 100) %>%
  ungroup()->
  rotation_severity_landscapes

length <- sum((standlevel_dist_events_joined %>% group_by(plotid) %>% summarize(ch_length = 1990 - min(year)))$ch_length)

standlevel_dist_events_joined %>% 
  select(plotid, year, severity) %>% 
  mutate(severityclass = floor(severity/5) * 5) %>%
  group_by(severityclass) %>% 
  summarize(distevents_n = n()) %>%
  arrange(desc(severityclass))%>%
  mutate(rotation_s = length/cumsum(distevents_n)) %>%
  select(-distevents_n) ->
  rotation_severity
```

## Rotation period of disturbance patch size

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
dist_patches %>% 
  group_by(newstand) %>% 
  summarize(nevents = n()) %>%
  slice(rep(1:n(), each = 1000)) %>%
  ungroup() %>% 
  mutate(rep = rep(1:1000, times = nrow(.) / 1000)) ->
  rotation_patchsize_boot

rotation_patchsize_boot %>%
  group_by(rep) %>%
  sample_n(., size = 20, replace = TRUE) %>%
  left_join(., dist_patches, by = "newstand") %>%
  mutate(patchsizeclass = floor(patch_area/5) * 5) %>%
  group_by(newstand, rep) %>% 
  mutate(ch_length = 1990 - min(peakyear)) ->
  rotation_patchsize_boot

rotation_patchsize_boot %>% 
  group_by(rep, newstand) %>% 
  summarize(ch_length = mean(ch_length)) %>%
  group_by(rep) %>% 
  summarize(rep_length = sum(ch_length)) ->
  rep_length

rotation_patchsize_boot %>% 
  group_by(rep, patchsizeclass) %>%
  summarize(distevents_n = n()) %>%
  group_by(rep) %>%
  complete(patchsizeclass = seq(0, 60, 5), fill = list(distevents_n = 0)) %>%
  left_join(., rep_length, by = "rep") %>%
  arrange(rep, desc(patchsizeclass)) %>%
  group_by(rep) %>%
  mutate(rotation_s = rep_length/cumsum(distevents_n)) %>%
  group_by(patchsizeclass) %>% 
  summarize(conf95 = quantile(rotation_s, probs = 0.975),
            conf05 = quantile(rotation_s, probs = 0.025)) ->
  rotation_patchsize_confidence

dist_patches %>%
  left_join(., 
            plots_clean %>% 
              filter(!is.na(newstand)) %>%
              group_by(locality, country, landscape, newstand) %>%
              summarize(nplots = n()),
            by = "newstand") %>%
  group_by(landscape, newstand) %>% 
  summarize(ch_length = 1990 - min(peakyear)) %>%
  group_by(landscape) %>% 
  summarize(lengthp = sum(ch_length))->
  length_patch_landscapes

dist_patches %>% 
  select(newstand, peakyear, patch_area) %>% 
  mutate(patchsizeclass = floor(patch_area/5) * 5) %>%
  left_join(., 
            plots_clean %>% 
              filter(!is.na(newstand)) %>%
              group_by(locality, country, landscape, newstand) %>%
              summarize(nplots = n()), 
            by = "newstand") %>%
  group_by(landscape, patchsizeclass) %>%
  summarize(distevents_n = n()) %>%
  complete(patchsizeclass = seq(0, 60, 5), fill = list(distevents_n = 0)) %>%
  arrange(landscape, desc(patchsizeclass)) %>%
  left_join(., length_patch_landscapes, by = "landscape") %>%
  mutate(rotation_patch = lengthp/cumsum(distevents_n)) %>%
  select(-distevents_n) %>%
  ungroup()->
  rotation_patchsize_landscapes

length_patch <- sum((dist_patches %>% group_by(newstand) %>% summarize(ch_length = 1990 - min(peakyear)))$ch_length)

dist_patches %>% 
  select(newstand, peakyear, patch_area) %>% 
  mutate(patchsizeclass = floor(patch_area/5) * 5) %>%
  group_by(patchsizeclass) %>% 
  summarize(distevents_n = n()) %>%
  arrange(desc(patchsizeclass))%>%
  mutate(rotation_s = length_patch/cumsum(distevents_n)) %>%
  select(-distevents_n) ->
  rotation_patchsize
```

## Rotation period of stand proportion disturbed

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
plotsprop_disturb %>% 
  group_by(newstand) %>% 
  summarize(nevents = n()) %>%
  slice(rep(1:n(), each = 1000)) %>%
  ungroup() %>% 
  mutate(rep = rep(1:1000, times = nrow(.) / 1000)) ->
  rotation_prop_boot

rotation_prop_boot %>%
  group_by(rep) %>%
  sample_n(., size = 20, replace = TRUE) %>%
  left_join(., plotsprop_disturb, by = "newstand") %>%
  mutate(proportionclass = floor(100 * plotsprop_disturb/5) * 5) %>%
  group_by(newstand, rep) %>% 
  mutate(ch_length = 1990 - min(peakyear)) ->
  rotation_prop_boot

rotation_prop_boot %>% 
  group_by(rep, newstand) %>% 
  summarize(ch_length = mean(ch_length)) %>%
  group_by(rep) %>% 
  summarize(rep_length = sum(ch_length)) ->
  rep_length

rotation_prop_boot %>% 
  group_by(rep, proportionclass) %>%
  summarize(distevents_n = n()) %>%
  group_by(rep) %>%
  complete(proportionclass = seq(0, 125, 5), fill = list(distevents_n = 0)) %>%
  left_join(., rep_length, by = "rep") %>%
  arrange(rep, desc(proportionclass)) %>%
  group_by(rep) %>%
  mutate(rotation_s = rep_length/cumsum(distevents_n)) %>%
  group_by(proportionclass) %>% 
  summarize(conf95 = quantile(rotation_s, probs = 0.975),
            conf05 = quantile(rotation_s, probs = 0.025)) ->
  rotation_prop_confidence

plotsprop_disturb %>%
  group_by(landscape, newstand) %>% 
  summarize(ch_length = 1990 - min(peakyear)) %>%
  group_by(landscape) %>% 
  summarize(lengthpr = sum(ch_length))->
  length_prop_landscapes

plotsprop_disturb %>% 
  select(newstand, landscape, peakyear, plotsprop_disturb) %>%
  mutate(proportionclass = floor(100 * plotsprop_disturb/5) * 5) %>%
  group_by(landscape, proportionclass) %>% 
  summarize(distevents_n = n()) %>%
  complete(proportionclass = seq(0, 125, 5), fill = list(distevents_n = 0)) %>%
  arrange(landscape, desc(proportionclass))%>%
  left_join(., length_prop_landscapes, by = "landscape") %>%
  mutate(rotation_prop = lengthpr/cumsum(distevents_n)) %>%
  select(-distevents_n) %>% 
  filter(proportionclass < 105) %>%
  ungroup()->
  rotation_prop_landscapes

length_prop <- sum((plotsprop_disturb %>% group_by(newstand) %>% summarize(ch_length = 1990 - min(peakyear)))$ch_length)

plotsprop_disturb %>% 
  select(newstand, peakyear, plotsprop_disturb) %>%
  mutate(proportionclass = floor(100 * plotsprop_disturb/5) * 5) %>%
  group_by(proportionclass) %>% 
  summarize(distevents_n = n()) %>%
  arrange(desc(proportionclass))%>%
  mutate(rotation_s=length_prop/cumsum(distevents_n)) %>%
  select(-distevents_n) ->
  rotation_prop
```
